Using Camera Traps and Computer Vision to Quantify Wildlife Diversity and Co-occurrence Where the Hudsons's Bay Lowlands meets the Ontario Shield

### 2024-04-08 reports/GnC_report/paper/paper.qmd
quick update
working through comments

---
title: "Using Camera Traps and Computer Vision to Quantify Wildlife Diversity and Co-Occurrence across \nOntario’s Far North"
subtitle: "Prepared for: Canadian Wildlife Service, Canadian Forest Service, and Wilfrid Laurier University"
author:
  - name: Valentin Lucet
    email: valentin.lucet@gmail.com
    orcid: 0000-0003-0268-818X
    affiliation: 
      - ref: wlu
  - name: Samantha McFarlane
    affiliation:
      - ref: eccc
  - name: Jennifer Baltzer
    affiliation:
      - ref: wlu
  - name: Cheryl A. Johnson
    affiliation:
      - ref: eccc
  - name: Eric Neilson
    affiliation:
      - ref: nrcan
  - name: Philip Weibe
    affiliation:
      - ref: nrcan
  - name: Frances Stewart
    orcid: 0000-0001-9344-8346
    affiliation:
      - ref: wlu
  - name: Josie Hughes
    affiliation:
      - ref: eccc
affiliations:
  - id: wlu
    name: Wilfrid Laurier University
  - id: eccc
    name: Environment & Climate Change Canada
  - id: nrcan
    name: Natural Resources Canada
title-block-published: "Last updated"  
date: now
date-format: long
format:
  html: default
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    extra_dependencies: ["float"]
format-links: false
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
bibliography: references.bib
abstract: >
  Northern Ontario holds some of the world’s last intact wild places, but it is under pressure from resource extraction and climate change. Our understanding of biodiversity measures like species occurrence, abundance, diversity, and density for the region are lacking, yet needed for strategic decision-making about the impacts of resource extraction on ecosystem services and biodiversity. The use of non-invasive survey methods, such as camera trap arrays, has significantly changed our ability to quantify wildlife across large and remote regions. However, these methods generate a large amount of data and require tools to speed up the identification of species. Machine learning methods, such as deep neural networks, are a promising tool to overcome this data bottle-neck. Yet, variables like strong seasonality, changes in vegetation type, and species characteristics can strongly impact their efficacy. The extent to which these methods perform for northern regions is not well established. Using an array of cameras deployed across an heterogeneous sampling area straddling two Northern Ontario ecozones (Hudson Bay Lowlands and Ontario Shield), we first look to quantify current species diversity and co-occurrence. We’ve recently developed an open-source workflow for camera trap image identification, access, and storage. We continue to use this workflow for ongoing image identification while investigating requirements to develop a machine learning image classifier.

@@

## Introduction

@@

Machine learning methods, such as deep neural networks, are now in use to automate the process of species identification and help in that endeavor. We aim to evaluate the viability of deep neural networks for camera trap surveys currently taking place in Northern Ontario’s heterogeneous landscapes, including the Hudson Bay Lowlands and Ontario Shield ecozones, as variables like weather, seasonality, vegetation type, and species characteristics can impact the performance of these methods. The extent to which these methods perform for northern regions is not well established. There are currently no available machine learning models for identifying species from images in northern regions and developing one would be beneficial as a monitoring and modeling tool for northern boreal ecosystems.

@@

## Methods

```{r echo=FALSE}
#| label: load-data
anns_wide <- readRDS(here::here("data", "objects", "annotations_wide_treated.rds"))
anns_wide_noempty  <- readRDS(here::here("data", "objects", "annotations_wide_noempty.rds"))
dets <- readRDS(here::here("data", "objects", "detections.rds"))
exif <- readr::read_csv(here::here("data", "tabular", "exif_processed.csv"))
cams_sf_coords <- readRDS(here::here("data", "objects", "cams_sf_coords.rds"))
zones_filtered  <- readRDS(here::here("data", "objects", "zones_filtered.rds"))
zones_pols <- readRDS(here::here("data", "objects", "zones_pols.rds"))
base_map <- readRDS(here::here("data", "objects", "base_map.rds"))
camera_days <- readRDS(here::here("data", "objects" ,"camera_days.rds"))
camera_regions <- readRDS(here::here("data", "objects" ,"camera_regions.rds"))
extr_table_sub_tb <- readRDS(here::here("data", "objects" ,"extr_table_sub_tb.rds"))
anns_wide_animals_joined_mut <- readRDS(here::here("data", "objects", "annotations_animals.rds"))
anns_wide_animals_table_tb <- readRDS(here::here("data", "objects" ,"anns_wide_animals_table_tb.rds"))
events_tb_filt_with_col <- readRDS(here::here("data", "objects" ,"events_tb_filt_with_col.rds"))
events_cam_days <- readRDS(here::here("data", "objects" ,"events_cam_days.rds"))
events_tb_cam <- readRDS(here::here("data", "objects" ,"events_tb_cam.rds"))
sci_names_processed <- readRDS(here::here("data", "objects" ,"sci_names_processed.rds"))
table_cat <- readRDS(here::here("data", "objects" ,"table_cat.rds"))
```

### Camera deployment and setup

@@

Briefly, 5 ARUs are deployed per site, with 100 m between ARUs. Up to 4 cameras were paired with ARUs at each of these sites, with at least one of these cameras placed at a random ARU site and one other prioritized to focus on a game trail or open area to increase detection probability (in total, we have 43 plots with 4 cameras, 5 plots with 3 cameras and 3 plots with 2 cameras). @fig-deployment shows the location of the clustered cameras and their current retrieval status. 170 cameras were retrieved in September 2022, totaling a deployment spanning 5-7 months. 5 additional cameras were only retrieved in March 2023 due to logistical and inaccessibility issues. There remains 18 cameras in the field to be retrieved in 2023.

The work conducted under this project focuses on quantifying and developing an open-source workflow to quantifying the wildlife species detected in those images that have been retrieved to date.

@@

regions_labels <- zones_filtered |> 
  sf::st_drop_geometry() |> 
  dplyr::bind_cols(data.frame(x = c(-84.5, -88.5),
                              y = c(52, 52.6)))
```

```{r echo=FALSE}
#| label: fig-deployment
#| fig-cap: "Camera deployment status for the 193 cameras, 170 of which were retrieved in September 2022. Labels for the 3 ecoregions spanned by the sampling area are included."

p <- base_map +
  ggplot2::geom_label(ggplot2::aes(x = x, y = y, label = ECOZONE_NA), inherit.aes = FALSE,
                      size = 2.5, data = regions_labels) +
  ggplot2::geom_point(data = cams_sf_coords,
                      shape = 24, col = "black", fill = "red",
                      ggplot2::aes(x = X, y = Y), 
                      inherit.aes = FALSE)

@@

p
```

### Image detection

@@

Images were retrieved from all cameras using a 32 GB SD card and copied onto a 10 TB hard drive (LaCie d2 Professional). The images amounted to 1.8TB of memory on this disk, for a total of **`r nb_image_fmt`** images. Images were stored with the following directory structure:

The images were processed through the animal detection model MegaDetector (@beery2019efficient), version 5.0, using a Lenovo X1 laptop with a NVIDIA GeForce GTX 1650 Max-Q graphics card, which took about 2 weeks of quasi-continuous processing to complete. The model processes each image and produces “detections” (bounding boxes) at a variable confidence score.

@@

At this first step into the process, no filtering was applied beyond the model predictions which classifies images as either likely to contain an object of relevance or empty. About **`r dets_tb_percent[1]` %** of the images, (**`r dets_tb_fmt[1]`**) images, were found by the model to possibly contain either a person, animal, or vehicle. The model did not produce detections for **`r dets_tb_fmt[2]`** images, (**`r dets_tb_percent[2]` %**) of the total.

@@

Megadetector is an AI model which produces confidence scores with each of its predictive detections. With no filtering, the overwhelming majority of the detections are of very low confidence (85% of detections are below a score of 0.1). This explains the high number of "person" and "vehicle" detections which are of course unrealistic. It is recommended to filter low probability detections, and cut-off values of 0.1 or 0.2 are usually suggested in the Megadetector documentation.

Using a cut-off value for the detection confidence of 0.1 (,i.e. discarding all detections with confidence below 0.1), this left us with a total of **`r dets_tb_cat_sum_fmt`** detections to sort through (see @fig-detections-histogram for an histogram of the detection scores, grouped by detection types). After applying this confidence threshold (decided on by the Wildlife Working Group) the total images to process for each category are: **`r dets_tb_cat_fmt[1]`** animals (**`r dets_tb_cat_percent[1]` %**); **`r dets_tb_cat_fmt[2]`** persons (**`r dets_tb_cat_percent[2]` %**); and **`r dets_tb_cat_fmt[3]`** vehicles (**`r dets_tb_cat_percent[3]` %**).

```{r echo=FALSE}
#| label: fig-detections-histogram
#| fig-cap: "Histogram of detection confidence scores produced for each bounding box predicted by Megadetector, showing a bimodal distribution. The colors represent the three categories of object identified by Megadetector: animals, persons and vehicles. Note that the hsitogram is truncated at our selected confidence threshold of 0.1."
#| fig.width: 8
#| fig.height: 4.5

@@

Finally, it is important to note that not only pictures with detections were uploaded to the LabelStudio platform. As stated before, camera traps produce images in bursts (in our case, 5 images per trigger). It is possible for Megadetector to miss a detection within a burst, or to produce variable confidence scores within that burst, which could lead to missing true detections if not all pictures from a given burst are filtered in. Therefore, we included all the images from a given burst if any of the images in the burst include a detection with certainty above the threshold, including potentially empty images from that burst.

### Tagging platform selection

@@

-   *Deployable online*. This was to allow for multiple people to help process the images.
-   *Flexible tagging interface*. Few platforms allow for a full customization of the tagging interface, down to the smallest details.
-   *Ability to tag multiple bounding boxes per image*. Most platforms only allow for tagging the entire image, and not for portions of it (e.g. bounding boxes of detections produced by MegaDetector). This is necessary for when we will crop our images to train our neural network model.
-   *Optionally, open source,* to allow for maximum reproducibility of the process.

To guide our choice, we used these criteria to informally compare 51 different platforms and tools currently available for processing and managing camera trap images. We first looked at platforms traditionally used for tagging camera trap images, but found them often lacking in two main aspects: flexibility of the tagging interface, and ability to manipulate bounding box information.

These two features are often better addressed by tools not traditionally used for tagging camera trap images, but used by the larger community of machine learning practitioners. Indeed, machine learning datasets outside can be very diverse and often more detailed than camera trap datasets, requiring processing tools with more detailed features. Tagging is a common step in machine learning research and in industry, and if often referred to as labeling.

We found two platforms that matched all of our criteria: LabelStudio (@LabelStudio) and CVAT (@boris_sekachev_2020_4009388). After testing both platforms, LabelStudio was chosen because the data format it uses is more easily compatible with Megadetector's COCO format output, making it easier to setup. We deployed the Community edition LabelStudio (version 1.6.0rc5) on an Ubuntu virtual machine hosted by Compute Canada, using an Apache server and a NGINX reverse proxy, which is a standard approach for custom hosting of websites.

### Tagging progress

@@

Out of the **`r dets_tb_cat_fmt[1]`** animal detections, **`r processed_fmt`** have been processed to date using our platform (or **`r processed_pct` %**) for a total of **`r processed_img_fmt`** images (i.e. **`r processed_img_pct` %** of the total amount of images, or **`r processed_img_up_pct` %** of the total amount of images uploaded to the platform).

As stated above, all images of a given burst are viewed if at least one image in the burst contains a detections with confidence above the threshold. This include potentially empty images. If we add the empty images viewed as part of a burst, it brings the total amount to **`r viewed_fmt`** images, (i.e. **`r viewed_pct` %** of the total amount of images, or **`r viewed_pct_up` %** of the total amount of images uploaded to the platform).

@@

Out of the **`r processed_fmt`** detections, **`r dels_nb_fmt`** were false positives (or **`r dels_nb_pct` %**), where Megadetector identified something, but there was nothing. In addition, **`r non_mans_nb_fmt`** were true positives (or **`r non_mans_nb_pct` %**), to which **`r mans_nb_fmt`** false negatives were added (manually added bounding boxes, **`r mans_nb_pct` %**). Out of the **`r non_mans_nb_fmt`** true positives, **`r orig_tb_fmt[3]`** had to have their bounding box adjusted. The table below summarises information for the tagging progress.

@@

knitr::kable(sum_tab,
             caption = "Summary table on tagging progress")
```

The relatively high rate of false positives is attributable to a combination of factors. First, the high seasonality of Ontario's far north makes it so that it is difficult to predict whether a camera that was set up in the winter won't be overgrown by vegetation in the spring. Locations where vegetation has grown in front of the sensor, combined with windy events, creates a perfect storm for false cameras triggers when vegetation is moving in front of the sensors. In addition, Megadetector is a likely to mistake a certain proportion of those vegetation pictures for actual animals. We are currently exploring avenues to tackle this roadblock by identifying repeating bounding images and by using a model to sort false from true positives.

#### Reproducibility

We are committed to producing a reproducible workflow. All code necessary to reproduce this report is available at an online [repository](https://github.com/StewartWILDlab/rofcamtrap). The repo makes use of [**mdtools**](https://github.com/StewartWILDlab/mdtools) module, a python-based command line tool for converting Megadetector outputs into LabelStudio inputs, and LabelStudio outputs into CSV tables.

### Analysis

#### Species detections

We summarized our existing species data to date for analysis in two main ways:  (1) species presence/absence data for each camera and, (2) relative species abundance, measured as the number of days, within each month that each species was detected at a camera site. The detection time for species is highly variable.

@@

#### Ordination and clustering

We analysed the relative abundance data using well known numerical ecology community analysis methods (Legendre and Legendre 2013). As a first step, we combined clustering and ordination to visualize associations between species, camera sites, and land cover in the vicinity of the detection site. We used Ward clustering on the relative abundances matrix. We extracted land cover information in a 1-km radius around each camera site, using the Far North Land Cover data set from Ontario's Ministry of Natural Resources and Forestry. We computed the frequencies of land cover classes and used these as constraining variable in a redundancy analysis, a method well suited for multidimensional data.

## Data exploration and preliminary results

### Wildlife community compposition

>>>>

---
title: "Using Camera Traps and Computer Vision to Quantify Wildlife Diversity and Co-occurrence Where the Hudsons's Bay Lowlands meets the Ontario Shield"
subtitle: "Prepared for: Canadian Wildlife Service, Canadian Forest Service, and Wilfrid Laurier University - 2024 update"
author:
  - name: Valentin Lucet
    email: valentin.lucet@gmail.com
    orcid: 0000-0003-0268-818X
    affiliation: 
      - ref: wlu
  - name: Samantha McFarlane
    affiliation:
      - ref: eccc
  - name: Jennifer Baltzer
    affiliation:
      - ref: wlu
  - name: Cheryl A. Johnson
    affiliation:
      - ref: eccc
  - name: Eric Neilson
    affiliation:
      - ref: nrcan
  - name: Philip Weibe
    affiliation:
      - ref: nrcan
  - name: Frances Stewart
    orcid: 0000-0001-9344-8346
    affiliation:
      - ref: wlu
  - name: Josie Hughes
    affiliation:
      - ref: eccc
affiliations:
  - id: wlu
    name: Wilfrid Laurier University
  - id: eccc
    name: Environment & Climate Change Canada
  - id: nrcan
    name: Natural Resources Canada
title-block-published: "Last updated"  
date: now
date-format: long
format:
  # html: default
  pdf:
    toc: true
    number-sections: true
    number-offset: 1
    colorlinks: true
    extra_dependencies: ["float"]
format-links: false
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
bibliography: references.bib
abstract: >
  Northern Ontario holds some of the world’s last intact wild places, but it is under pressure from resource extraction and climate change. Our understanding of biodiversity measures like species occurrence, abundance, diversity, and density for the region are lacking, yet needed for strategic decision-making about the impacts of resource extraction on ecosystem services and biodiversity. The use of non-invasive survey methods, such as camera trap arrays, has significantly changed our ability to quantify wildlife across large and remote regions. However, these methods generate a large amount of data and require tools to speed up the identification of species. Machine learning methods, such as deep neural networks, are a promising tool to overcome this data bottle-neck. Yet, variables like strong seasonality, changes in vegetation type, and species characteristics can strongly impact their efficacy. The extent to which these methods perform for northern regions is not well established. Using an array of cameras deployed across an heterogeneous sampling area straddling two Northern Ontario ecozones (Hudson Bay Lowlands and Ontario Shield), we first look to quantify current species diversity and co-occurrence.In 2023 we developed an open source workflow for camera trap image identification, access, and storage. We have updated it in 2024 for efficiency. We also continue to investigate the requirements to develop a machine learning image classifier. {{< pagebreak >}}

@@

{{< pagebreak >}}

# Introduction

@@

Machine learning methods, such as deep neural networks, are now in use to automate the process of species identification and help in that endeavor. We aim to evaluate the viability of deep neural networks for camera trap surveys currently taking place in Northern Ontario’s heterogeneous landscapes, including the Hudson Bay Lowlands and Ontario Shield ecozones, as variables like weather, seasonality, vegetation type, and species characteristics can impact the performance of these methods. The extent to which these methods perform for northern regions and northern specie is not well established. There are currently no available machine learning models for identifying species from images in northern regions and developing one would be beneficial as a monitoring and modeling tool for northern boreal ecosystems.

@@

## Update for the 2024 progress report

Since the last progress report (2023 report), we flag the following improvements and changes, which are detailed in the new sections of this reports.

- The number of fully processed images was doubled from about 50000 to more than 100000, keeping with the rate of progress of the first year of tagging.
- We made a first attempt at at using machine learning methods to train a species classifier using about 23000 images as the basis dataset, which produced encouraging results.
- After reevaluation of the ability of our current worlflow and platform to scale to multiple years of data, we have transitioned the image tagging workflow to [WildTrax](https://wildtrax.ca/).
- This report repeatedly mentions the AI model MegaDetector. Although this model now falls under the toolset known as PytorchWildlife, as of writing in April 2024, it is still commonly known as MegaDetector by the camera trapping community.

### Outputs of note

1. Poster submitted and presented at 2023 conference of the Ontario chapter of The Wildlife Society (OCTWS).
2. Poster submitted and displayed at the Mushkegowuk council 2023 Research Meeting.
3. Application to participate in the Workshop on Computer Vision Methods for Ecology at the Caltech Resnick Sustainability Institute. Rejectec in 2023, reapplied in 2024.

# Methods

```{r echo=FALSE}
#| label: load-data
anns_wide <- readRDS(here::here("data", "objects", "annotations_wide_treated.rds"))
anns_wide_noempty  <- readRDS(here::here("data", "objects", "annotations_wide_noempty.rds"))
dets <- readRDS(here::here("data", "objects", "detections.rds"))
exif <- readr::read_csv(here::here("data", "tabular", "exif_processed.csv"))
cams_sf_coords <- readRDS(here::here("data", "objects", "cams_sf_coords.rds"))
zones_filtered  <- readRDS(here::here("data", "objects", "zones_filtered.rds"))
zones_pols <- readRDS(here::here("data", "objects", "zones_pols.rds"))
base_map <- readRDS(here::here("data", "objects", "base_map.rds"))
ont_base_map <- readRDS(here::here("data", "objects", "ont_base_map.rds"))
camera_days <- readRDS(here::here("data", "objects" ,"camera_days.rds"))
camera_regions <- readRDS(here::here("data", "objects" ,"camera_regions.rds"))
extr_table_sub_tb <- readRDS(here::here("data", "objects" ,"extr_table_sub_tb.rds"))
anns_wide_animals_joined_mut <- readRDS(here::here("data", "objects", "annotations_animals.rds"))
anns_wide_animals_table_tb <- readRDS(here::here("data", "objects" ,"anns_wide_animals_table_tb.rds"))
events_tb_filt_with_col <- readRDS(here::here("data", "objects" ,"events_tb_filt_with_col.rds"))
events_cam_days <- readRDS(here::here("data", "objects" ,"events_cam_days.rds"))
events_tb_cam <- readRDS(here::here("data", "objects" ,"events_tb_cam.rds"))
sci_names_processed <- readRDS(here::here("data", "objects" ,"sci_names_processed.rds"))
table_cat <- readRDS(here::here("data", "objects" ,"table_cat.rds"))
```

## Camera deployment methods

@@

TODO: add more from Sam

## Deployment summary

Briefly, 5 ARUs are deployed per site, with 100 m between ARUs. Up to 4 cameras were paired with ARUs at each of these sites, with at least one of these cameras placed at a random ARU site and one other prioritized to focus on a game trail or open area to increase detection probability. Weather restrictions and field conditions limited access to some sites and resulting in some variation in this design. In total we have 43 plots with 4 cameras, 5 plots with 3 cameras and 3 plots with 2 cameras. @fig-deployment shows the location of the clustered cameras and their current retrieval status. 54 cameras were retrieved in May 2022 and 116 in September 2022. 6 additional cameras were retrieved in March 2023 due to logistical and accessibility issues. Another 77 cameras were retrieved in Mar 2023. The last 18 cameras were retrieved in September 2023 and are now being processed.

The work conducted under this project focuses on developing an open-source workflow to quantify the wildlife species detected in those images that have been retrieved to date. This workflow is also being designed to include anticipated additional images from future deployments on this, and potentially other, projects.

@@

regions_labels <- zones_filtered |> 
  dplyr::mutate(ECOZONE_NA = c("Hudson Bay\nLowlands", "Ontario\nShield")) |> 
  sf::st_drop_geometry() |> 
  dplyr::bind_cols(data.frame(x = c(-84.5, -88.75),
                              y = c(52, 52.6)))
```

```{r echo=FALSE}
#| label: fig-deployment
#| fig-cap: "Left: Map of the province of Ontario with ecoregion and inset map delimitation. Right: Inset map of camera deployment locations as wellas deployment and retrieval dates for the 227 sites. Labels for the 2 ecoregions spanned by the sampling area are included."
#| fig.width: 8
#| fig.height: 6

p <- base_map +
  ggplot2::geom_label(ggplot2::aes(x = x, y = y, label = ECOZONE_NA), inherit.aes = FALSE,
                      size = 2.5, data = regions_labels) +
  ggplot2::geom_point(data = cams_sf_coords,
                      shape = 24, col = "black", #fill = "red",
                      ggplot2::aes(x = X, y = Y, fill = deployment_group), 
                      inherit.aes = FALSE) +
  ggplot2::labs(fill = "Deployment - retrieval") #+
  # ggplot2::theme_void() +
  # ggplot2::theme(
  #   # legend.justification defines the edge of the legend that the legend.position coordinates refer to
  #   legend.justification = c(0, 1),
  #   # Set the legend flush with the left side of the plot, and just slightly below the top of the plot
  #   legend.position = c(0, .95)
  #   ) +
  # ggplot2::theme(
  #   text = ggplot2::element_text(family = "Futura-Medium"),
  #   legend.title = ggplot2::element_text(family = "Futura-Bold", size = 10),
  #   legend.text = ggplot2::element_text(family = "Futura-Medium", size = 10)
  #   ) +

@@

cowplot::plot_grid(ont_base_map, p, ncol = 2, rel_widths = c(1,2))
```

## Data access and storage

Images were retrieved from all cameras using a 32 GB SD card and copied onto a 10 TB hard drive (LaCie d2 Professional). The images amounted to 2+TB of memory on this disk, for a total of **`r nb_image_fmt`** images. They were stored with the following directory structure:

We have made a backup of the these images on a compute Canada cloud instance , accessible via SSH terminal. We also plan to keep an additional copy of the images on a separate harddrive.

## Image processing

In this section, we separate the workflow we developed and used between September 2022 and December 2023, from the new workflow we are experimenting with and hoping to perfect for 2024 and onward. In 2022 we explored alternative tagging methods as no platform was at that time equipped for our needs. We developed a custom and reproducible platform for image tagging through lable studio and R scripts. We deployed our own LabelStudio platform and used custom (yet reproducible) scripts for image analysis. This has proven to be a workable yet flawed method for processing our large amount of images. The main issue we ran into is due to the fact that MegaDetector (which is described in more details in the next section), produced large amount of false positives in our region. This may have resulted from Megadetector not being calibrated to highly seasonal landscapes, resulting in very slow processing of images through Labelstudio. This made data processing in LabelStudio slower and more cumbersome than we initially anticipated, as LabelStudio is not well equipped for group tagging of false positives. However, in the absence of a better alternative, we persevered in using the platform, and attempted to circumvent these issues by trying to reduce the number of false positives in the MegaDetector output. We attempted to find patterns in the false positives rates and to train a separate classifier but we were unsuccessful. As we continued, it slowly became clear that LabelStudio could not scale appropriately to multi-year data. Concurrently, the wildtrax platform improved significantly. Wildtrax now supports MegaDetector and image filtering based on score; something we had been conducting manually through our LabelStudio protocol. It started being able to display the bounding boxes created by MegaDetector and exposed these results in its API, allowing for further processing. Compared to 2022, it now has much better batch tagging and species verification abilities. Currently, it is still missing the ability to assign specific tags to bounding boxes within the app. As of March 2024 we are actively shifting out workflow to WildTrax, reanalyzing all previous work conducted in LabelStudio. For transparency we outline both LabelStudio and wildtrax protocols below.

### Previous workflow (LabelStudio)

#### Image detection

@@

For the LabelStudio workflow, the images were first processed through the animal detection model MegaDetector (@beery2019efficient), version 5.0, using a Lenovo X1 laptop with a NVIDIA GeForce GTX 1650 Max-Q graphics card, which took about 2 weeks of quasi-continuous processing to complete. MegaDetector processes each image and produces “detections” (bounding boxes) at a variable confidence score.

@@

At this first step into the process, no filtering was applied beyond the model predictions which classifies images as either likely to contain an object of relevance or empty. About **`r dets_tb_percent[1]` %** of the images, (**`r dets_tb_fmt[1]`** images), were found by the model to possibly contain either a person, animal, or vehicle. The model did not produce detections for **`r dets_tb_fmt[2]`** images, (**`r dets_tb_percent[2]` %**) of the total.

@@

MegaDetector is an AI model which produces confidence scores with each of its predictive detections. With no filtering, the overwhelming majority of the detections are of very low confidence (85% of detections are below a score of 0.1). This explains the high number of "person" and "vehicle" detections which are of course unrealistic. It is recommended to filter low probability detections, and cut-off values of 0.1 or 0.2 were suggested in the [MegaDetector documentation](https://github.com/microsoft/CameraTraps) at the time model was ran.

As the NSERC Alliance Wildlife Working Group, we decided to use a cut-off value of 0.1 for detection confidence of 0.1 (,i.e. discarding all detections with confidence below 0.1), leaving us with a total of **`r dets_tb_cat_sum_fmt`** detections to sort through (see @fig-detections-histogram for an histogram of the detection scores, grouped by detection types). After applying this confidence threshold (decided on by the Wildlife Working Group) the total images to process for each category were: **`r dets_tb_cat_fmt[1]`** animals (**`r dets_tb_cat_percent[1]` %**); **`r dets_tb_cat_fmt[2]`** persons (**`r dets_tb_cat_percent[2]` %**); and **`r dets_tb_cat_fmt[3]`** vehicles (**`r dets_tb_cat_percent[3]` %**).

```{r echo=FALSE}
#| label: fig-detections-histogram
#| fig-cap: "Histogram of detection confidence scores produced for each bounding box predicted by MegaDetector on camera trap images collected from the Northern Ontario array between March 2022 and March 2023, showing a bimodal distribution. The colors represent the three categories of object identified by MegaDetector: animals, persons and vehicles. Note that the hsitogram is truncated at our selected confidence threshold of 0.1."
#| fig.width: 8
#| fig.height: 4.5

@@

Finally, it is important to note that not only pictures with detections were uploaded to the LabelStudio platform. Camera traps produce images in bursts (in our case, 5 images per trigger). It is possible for MegaDetector to miss a detection within a burst, or to produce variable confidence scores within that burst, which could lead to missing true detections if not all pictures from a given burst are included. Therefore, if any image from a burst exceeded the selected MegaDetector threshold, we included it in our analysis .

#### Tagging platform selection

@@

-   *Deployable online*. This was to allow for multiple people to help process the images.
-   *Flexible tagging interface*. Few platforms allow for a full customization of the tagging interface, down to the smallest details.
-   *Ability to tag multiple bounding boxes per image*. Most platforms only allow for tagging the entire image, and not for portions of it (e.g. bounding boxes of detections produced by MegaDetector). This is necessary for when we will crop our images to train our neural network model.
-   *Optionally, open source,* to allow for maximum reproducibility of the process.
-   *Ideally,* stored on a cloud system hosted in Canada.

To guide our choice, we used these criteria to informally compare 51 different platforms and tools currently available for processing and managing camera trap images (a list of the tools compared in given in Appendix III). We first looked at platforms traditionally used for tagging camera trap images, but found them often lacking in two main aspects: flexibility of the tagging interface, and ability to manipulate bounding box information.

These two features are often better addressed by tools not traditionally used for tagging camera trap images, but used by the larger community of machine learning practitioners. Indeed, machine learning datasets can be very diverse and often more detailed than camera trap datasets, requiring processing tools with more detailed features. Tagging is a common step in machine learning research and in industry, and if often referred to as labeling.

We found two platforms that matched all of our criteria: LabelStudio (@LabelStudio) and CVAT (@boris_sekachev_2020_4009388). After testing both platforms, LabelStudio was chosen because the data format it uses is more easily compatible with MegaDetector's COCO format output, making it easier to setup. We deployed the Community edition LabelStudio (version 1.6.0rc5) on an Ubuntu virtual machine hosted by Compute Canada, using an Apache server and a NGINX reverse proxy, which is a standard approach for custom hosting of websites.

### Updated workflow (Wildtrax)

#### Image renaming and organization

In order to reduce the impact of false positives on our workflow, we resorted to adding a pre-processing step in our workflow in which all images were manually sorted into "wildlife" and "nonwildlife" folders inside each subfolders. In addition, because wildtrax does not track the upload path of an image, and instead tracks the unique name of each file, we are required to change the way we encode unique images. Therefore, we developed a simple and very efficient script in the Julia programming language to paralellize the renaming of each file into a standardized manner. We had originally refrained from using this method of uniquely identifying files and instead relied on tracking the unique path of each file as we wanted to reduce the impact on our storage capacity (as it essentially required to double the required storage space).

The standard format we used for file renaming follows this convention:

- "\textcolor{red}{retrievalID}\_\textcolor{blue}{siteID}\_\textcolor{green}{locationID}\_\textcolor{purple}{cameraID}\_\textcolor{orange}{subfolderID}\_\textcolor{PineGreen}{fileID}\_\textcolor{cyan}{dateTime}.JPG",

Giving, for example, a file with the name:

- "\textcolor{red}{TC1}\_\textcolor{blue}{P028}\_\textcolor{green}{P028-1}\_\textcolor{purple}{WLU-10}\_\textcolor{orange}{100}\_\textcolor{PineGreen}{0001}\_\textcolor{cyan}{2022\_03\_18\_09\_44\_10}.JPG".

We also added a "W" if the image had been sorted into a "wildlife" folder, ie: "TC2_INTC02_INTC02-2_CWS-ON46_*W*_0099_2022_12_17_13_41_00.JPG".

#### Current tagging workflow and postprocessing

Wildtrax keeps track of image batches using "location folders". We created our own location folders but combining images within the same retrieval, site and location IDs. For example the image mentioned above would be placed in the "TC1_P028_P028-1_WLU-10" location folders. For locations in which images were previously sorted into "wildlife" and "nonwildlife", we declined each location folders into a "_W" and "_NW" variants and uploaded as such to wildtrax.

As a first step in testing the wildtrax workflow, we are uploading the folders from the first retrieval that showed the highest rates of false positives, retagging and evaluating the level of overlap with the LabelStudio workflow and also allowing us to complete these folders. As a second step, we are uploading the folders from the very last retrieval, to speed up the tagging progress. What remains to be determined is how we will properly assign tags to bounding boxes. We are currently considering two main avenues: 1. re-purposing the previous LabelStudio workflow for bounding box assignment, and 2. using a non networked platform, for example the EcoAssist software.

## Species Classifier

We made a first attempt at developing a species classifier using the 20000+ cropped bounding boxes resulting from the tagging process. We used the Pytorch framework, training a resnet100 model for 100 epochs, and modifying the last layer to include a regularization (i.e. dropout) layer in order to avoid overfitting. We tested the species classifier training via both transfer learning (frozen weights) and fresh learning (unfrozen weights). We used a batch size of 32.

## Tagging progress - LabelStudio platform

@@

Out of the **`r dets_tb_cat_fmt[1]`** animal detections, **`r processed_fmt`** have been processed to date using the LabelStudio platform (or **`r processed_pct` %**) for a total of **`r processed_img_fmt`** images (i.e. **`r processed_img_pct` %** of the total amount of images, or **`r processed_img_up_pct` %** of the total amount of images uploaded to the platform).

As stated above, all images of a given burst are viewed if at least one image in the burst contains at least one detection with confidence above the MegaDetector confidence threshold. This includes potentially empty images. If we add the empty images viewed as part of a burst, it brings down the total amount of viewed images to **`r viewed_fmt`** images, (i.e. **`r viewed_pct` %** of the total amount of images, or **`r viewed_pct_up` %** of the total amount of images uploaded to the platform).

@@

Out of the **`r processed_fmt`** detections, **`r dels_nb_fmt`** were false positives (or **`r dels_nb_pct` %**). In addition, **`r non_mans_nb_fmt`** were true positives (or **`r non_mans_nb_pct` %**), to which **`r mans_nb_fmt`** false negatives were added (manually added bounding boxes, **`r mans_nb_pct` %**). Out of the **`r non_mans_nb_fmt`** true positives, **`r orig_tb_fmt[3]`** had to have their bounding box adjusted: the bounding box drawn by MegaDetector was at least partially innacurate and had to be corrected. The table below summarises information for the tagging progress.

@@

knitr::kable(sum_tab,
             caption = "Summary table on tagging progress. The 0.1 filter value corresponds to the megadetector cinfidence threshold.")
```

The relatively high rate of false positives is attributable to a combination of factors. First, the high seasonality of Ontario's far north makes it difficult to predict whether a camera that was set up in the winter won't be overgrown by vegetation in the spring. This isn't a unique problem to this project, but an issue in many ecosystems. Locations where vegetation has grown in front of the sensor, combined with windy events, increases the number of photos triggered by vegetation, rather than wildlife. MegaDetector is likely to mistake many of the vegetation pictures for animals. We are currently exploring avenues to ways to address this by identifying repeating bounding images and by using a model to sort false from true positives.

## Tagging progress - WildTrax

TODO: update once data is harmonized

## Reproducibility

We are committed to producing a reproducible workflow. All code necessary to reproduce this report is available at an online [repository](https://github.com/StewartWILDlab/rofcamtrap), which is currently private but will be opened upon submission of the final version of this report. The repo makes use of the [**mdtools**](https://github.com/StewartWILDlab/mdtools) module, a python-based command line tool for converting MegaDetector outputs into LabelStudio inputs, and LabelStudio outputs into CSV tables.

## Analysis

### Species detections

We summarized our existing species data in two ways:  (1) species presence/absence at each camera and, (2) relative species abundance, measured as the number of days, within each month that each species was detected at a camera site. The detection time for species is highly variable.

@@

### Ordination and clustering

We analysed the relative abundance data using published and standardized numerical ecology community analysis methods (Legendre and Legendre 2013). As a first step, we combined clustering and ordination to visualize associations between species, camera sites, and land cover in the vicinity of the detection site. We used Ward clustering on the relative abundances matrix. We extracted land cover information in a 1-km radius around each camera site, using the Far North Land Cover data set from Ontario's Ministry of Natural Resources and Forestry. We computed the frequencies of land cover classes and used these as constraining variable in a redundancy analysis, a method well suited for multidimensional data.

# Data exploration and preliminary results

## Wildlife community composition

We obtained repeat detections of at least 31 mammal and bird species across 193 sites and 5-7 months of deployment. The most common species detected (detected more than 50 days in the season) are listed in Table 2 below (the full table is available in Appendix I). Note the large amount detections in the “bird” and “duck” species categories which could be further distinguished into specific species; eight of the top 17 detected species categories were avian.
